---
title: "Assignment 1 - Language Development in ASD - part 2"
author: "Esther Dyngby Jensen"
date: "September 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Language development in Autism Spectrum Disorder (ASD)

Background: Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail: i) relying on actual naturalistic language production, ii) over extended periods of time. We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 
NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

We then want to test the language trajectory of child and parent over time.

This Markdown is structured in the following way:

1. The exercises: read them carefully. Under each exercise you will have to write your answers, once you have written and run the code. This is the part that you have to directly send to the teachers.
2. An (optional) guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results under the exercise part.
3. In exercise 4 you will be asked to create the best possible model of language development in TD and ASD children, picking and choosing whatever additional variables you want from the dataset. Next time, the models produced by the different groups will compete against each other to see who can produce the best model, so choose carefully!

You will have to have a github repository for the code and send the answers to Celine and Riccardo without code (but a link to your github repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-) 

N.B. The following lines are a summary of the questions to be answered, the step-by-step instructions and tips are below.

## Exercise 1) Preliminary Data Exploration

Describe the participant samples in the dataset (e.g. by diagnosis, age, etc.). Do you think the two groups are well balanced? If not, what do you think was the reason?

Testing different parameters on typically developed childen and ASD children in the dataset, it appears that there is a significant difference between age, as it seems ASD childeren are significantly older than the TD children (b=-12.63 , se = 1.02 , t-value(60) = -12.32 , p<0.001)
The reason for choosing that the two groups are that different in reference to age might be for the purpose of making the two groups more comparable. The remaining parameters measured(gender, non-verbal IQ and verbal IQ) do not differ significantly in between groups.

### Exercise 2) Children learning language: the effects of time and ASD
Describe linguistic development in TD and ASD children in terms of Mean Length of Utterance (MLU)?

[HERE GOES YOUR ANSWER]

### Exercise 3) Child directed speech as a moving target
Describe how parental use of language changes over time in terms of MLU. What do you think is going on?

MLU appears to increase over time across groups of diagnosis. An overall increase as well as an increase in children MLU, may indicate as the children talk more, so does the parent. Parent MLU changes significantly over time (b=.12, SE=0.01, t(60)=6.5 p<.001). There does seem to be a difference in terms of diagnosis as well. Parent of TD children have a significantly higher value of MLU than parents of ASD children (b=.5, SE=.11, t(60)=4.42, p<.001).

### Exercise 4) Looking into "individual differences" (demographic, clinical or cognitive profiles)
The dataset contains some additional variables characterizing the kids’ cognitive and clinical profile: ADOS (autism severity), MSEL EL (Expressive Language, that is, verbal IQ, or linguistic skills at first visit as assessed by a psychologist using Mullen Scales of Early Learning), MSEL VR (Visual Reception, used as a proxy for non verbal IQ at first visit), Age, Gender, Ethnicity. Would it make sense to add any of them to your model of linguistic trajectories? Create the best possible model (the one that best explain the data, with MLU as outcome). Next time your model will be tested on new participants, and we will proclaim a winner. Describe your strategy to select the best models (how did you choose the variables to include?) and send the code to Riccardo and Celine.

We attempted to find the variables which adds some sort of conceptual meaning, when predicting child MLU. As time has previously shown to be a significant predictor of MLU, this parameter is an obvious one to include. The amount of words spoken during a visit, also has an intuitive ring to it, as amount of words spoken also will increase the mean length of utterances. Verbal IQ was implemented in the model as the more complex sentences are formed by higher verbal IQ scores, and thus adding to the lenght of MLU. 

### [OPTIONAL] Exercise 5) Comment on how the three linguistic variables measure linguistic performance (the so-called "construct validity" of the measures). Do they express the same variance?

[HERE GOES YOUR ANSWER]

### Structure of the code chunks

Basic stuff:
- Loading the libraries
- Setting the directory and loading the data
- Look at the data (which variables are there? Are they in the right format?) and describe the participants (by diagnosis)

We will try to answer three questions:

- Do children with ASD develop language differently from non-ASD children?
- Do parents speak differently to children with ASD than to non-ASD ones?
- Which variables should we use to best explain the child linguistic performance?
  
### Loading the relevant libraries

Load necessary libraries : what will you need?

- e.g. something to plot with
- e.g. mixed effects models

```{r Load Libraries, message=FALSE}
library(lmerTest);library(lme4);library(ggplot2);library(tidyverse);library(MuMIn)

# Load Data 
setwd("~/OneDrive - Aarhus universitet/AU-Cognitive Science/3rd Semester/Experimental Methods 3/Exercise/Assignments/Assignment 1")
data=read.csv("clean_data.csv")
data=data[,-1]
```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Visit, Number of words used, Number of unique words used, length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r}
# Making a subset of 1st visit data
Visit1_data=subset(data, data$visit == 1)

#age test
modelAgeDiagnosis = lm(Age~Diagnosis,Visit1_data)
summary(modelAgeDiagnosis)

plot = ggplot(Visit1_data, aes(x = Diagnosis, y = Age)) + geom_violin() + ggtitle("Age and diagnosis")
plot

#gender test
modelG=glm(Gender~Diagnosis, family=binomial,Visit1_data)
summary(modelG)

#Non verbal IQ test
modelNVIQDiagnosis = lm(nonVerbalIQ~Diagnosis,Visit1_data)
summary(modelNVIQDiagnosis)

plot1 = ggplot(Visit1_data, aes(x = Diagnosis, y = nonVerbalIQ)) + geom_violin()+ ggtitle("Non-verbal IQ and diagnosis")
plot1

#Verbal IQ test
modelVIQDiagnosis = lm(verbalIQ~Diagnosis,Visit1_data)
summary(modelVIQDiagnosis)

plot2 = ggplot(Visit1_data, aes(x = Diagnosis, y = verbalIQ)) + geom_violin()+ggtitle("Verbal IQ and diagnosis")
plot2

```

#####Reporting results of models:
Age: there is a significant difference, ASD children are significantly older than TD children (b=-12.63 , se = 1.03 , t-value(60) = -12.32 , p<0.001).
Gender: results suggest that gender has a nonsignificant influence on the diagnosis (b=-0.36 , se = 0.7 , z-value(60) = -0.52  , p= 0.6)
Non-verbal IQ: non-verbal IQ causes no inequality between the two groups, as it appears the values for TD children are not significantly different from ASD children (b=-0.89 , se = 1.17 , t-value(60) = -0.76 , p=0.45).
Verbal IQ: TD children have no significantly higher verbal IQ than ASD children (b=2.9 , se = 1.625 , t-value(60) = 1.79 , p=0.78).

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r}
Scatterplot = ggplot(data, aes(visit, CHI_MLU, color=Diagnosis))+geom_point()+geom_smooth(method=lm)
Scatterplot

# Statistical test
mixedmodel=lmer(CHI_MLU~visit+Diagnosis +(1+visit|ID),data, REML=FALSE)
summary(mixedmodel)
```
The plot does not represent the model as each child doesn't have their individual intercept. Although is does work as a description of the data.

######Reporting model above
visit predicts MLU...
Diagnosis does not predict MLU

Fixed effects:visit and diagnosis. 
Random factors: ID and Visit.

####How would you evaluate whether the model is a good model?

```{r}
# Finding R-squared
r.squaredGLMM(mixedmodel)
#marginal//conditional The marginal R squared values are those associated with your fixed effects, the conditional ones are those of your fixed effects plus the random effects.


# Evaluating the model
# Question:
mixedmodel.null=lmer(CHI_MLU~visit + (1+visit|ID), data, REML=FALSE)
anova(mixedmodel.null,mixedmodel)
#not a good model, there's better ones, but then Diagnosis is omitted.
```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better

```{r}
#Riccardo: Growth curve model havent we talked about yet... We shall wait

Quadratic_m=lmer(CHI_MLU~visit+I(visit^2) + Diagnosis +(1+visit|ID),data, REML=FALSE)
Cubic_m=lmer(CHI_MLU~visit+I(visit^2)+I(visit^3) + Diagnosis +(1+visit|ID),data, REML=FALSE)
anova(mixedmodel,Quadratic_m,Cubic_m)

# qua is significantly better than the linear model(x^2(1)=28.87, p>.05), but the cubic model is not significantly better.

summary(Quadratic_m)

```

Exciting right?
Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r}
Scatterplot1 = ggplot(data, aes(visit, MOT_MLU, color=Diagnosis))+geom_point()+geom_smooth(method=lm)
Scatterplot1

mixedmodel1=lmer(MOT_MLU~visit+Diagnosis +(1+visit|ID),data, REML=FALSE)

summary(mixedmodel1)
r.squaredGLMM(mixedmodel1)

```

There is a significant difference in how the parental MLU changes over time, (b=.12, SE=0.01, t(60)=6.5 p<.001), as well as there appears to be a significant difference in parental MLU according to diagnosis, parents of TD children talked significantly more than the parents of ASD children (b=.5, SE=.11, t(60)=4.42, p<.001)

### Adding new variables (Exercise 4)
Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Celine.

```{r message=FALSE}
bestmodel=lmer(CHI_MLU~visit+MOT_MLU+verbalIQ+(1+visit|ID),data, REML=FALSE)
summary(bestmodel)
r.squaredGLMM(bestmodel)
```

When testing to find the best model, a dubious manner of experimental conduct was applied, while fishing for a high value of R2 and significant p-values. Other than that, the variables tested while fishing was only the ones which made good conceptual sense. 
